---
title: "Class 11 Lab Session"
author: "Garrett Cole"
format: pdf
---

# Section 1. Proportion on G/G in a population
```{r}
# Read in Mexican Ancestry in LA csv file
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

head(mxl)
```
```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100

```

## Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes
The corresponding median expression levels for each genotype is in the boxplot below
```{r}
#Read in file
expr <-  read.table("rs8067378_ENSG00000172057.6.txt")

head(expr)
```
```{r}
nrow(expr)

table(expr$geno)
```
```{r}
library(ggplot2)
ggplot(expr) + aes(geno, exp, fill = geno) +
  geom_boxplot(notch = TRUE)
```
## Generate a boxplot with a box per genotype, what could you infer from the relative  expression value between A/A and G/G displayed in this plot? Does the SNP effect the  expression of ORMDL3?
I could infer that A/A has a higher relative expression value than G/G while A/G has the most medium expression value. The SNP do affect the expression of ORMDL3
