---
title: "Class 17: Vaccination Mini-Project"
author: "Garrett Cole"
format: gfm
---

## Getting Started
```{r}
#import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```
# Q1: What column details the total number of people fully vaccinated?
10
```{r}
#Commented out for pdf space
#vax["persons_fully_vaccinated"]
#vax[10]
```
# Question 2: What column details the Zip code tabulation area? 
2
```{r}
#Commented out for pdf space
#vax["zip_code_tabulation_area"]
#vax[2]
```
```{r}
# dimensions of dataset
dim(vax)
```
# Question 3: What is the earliest date in this dataset?
2021-01-05
```{r}
vax$as_of_date[1]
```
# Question 4: What is the latest date in this dataset?
2022-11-22
```{r}
vax$as_of_date[174636]
```
```{r}
# Overview of vax
library(skimr)
skimr::skim(vax)
```
# Question 5: How many numeric columns are in this dataset?
There are 9 numeric columns
# Question 6: Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?
14921
```{r}
sum(is.na(vax$persons_fully_vaccinated))
```

# Question 7: What percent of persons_fully_vaccinated values are missing (to 2 significant figures)
8.54%
```{r}
#convert NA values to 0
vax[is.na(vax)] = 0
```
```{r}
#double check number of missing values is 14921
colSums(vax==0)
```
```{r}
# divide num. of missing values by total num. of values to find percent missing
(14921/174636) * 100
```
# Question 8 (Optional): Why might this data be missing?
This data might be missing because there wasn't a confirmed number of persons fully vaccinated for that entry at the time
## Working with Dates
```{r}
library(lubridate)
```
```{r}
today()
```
```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```
```{r}
# time since first date to today
today() - vax$as_of_date[1]
```
```{r}
# time since first date to last date
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```
# Question 9: How many days have passed since the last update of the dataset?
6 days since last update
```{r}
# time since last update to today's date
today() - vax$as_of_date[nrow(vax)]
```
# Question 10: How many unique dates are in the dataset (i.e. how many different dates are detailed)?
99 unique dates
```{r}
length(unique(vax$as_of_date))
```
## Working with Zip Codes
```{r}
library(zipcodeR)
```
```{r}
#centroid of La Jolla zip code 92037
geocode_zip('92037')
```
```{r}
#distance between two centroids
zip_distance('92037','92109')
```
```{r}
#census data about zip code areas
reverse_zipcode(c('92037', "92109") )
```
## Focus on San Diego Area
```{r}
# Subset to San Diego county only areas
sd <- vax[ vax$county == "San Diego" , ]
```
```{r}
#using dplyr
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```
```{r}
# all San Diego county areas with a population of over 10,000
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```
# Question 11: How many distinct zip codes are listed for San Diego County?
107 distinct zip codes
```{r}
length(unique(sd$zip_code_tabulation_area))
```
# Question 12: What San Diego County Zip code area has the largest 12 + Population in this dataset?
92154
```{r}
sd$zip_code_tabulation_area[which.max(sd$age12_plus_population)]
```
```{r}
#select all San Diego “county” entries on “as_of_date” “2022-11-15”
tempSD <- filter(vax, county == "San Diego" &
                as_of_date == "2022-11-15")
```

# Question 13: What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-11-15”?
0.6818
```{r}
mean(tempSD$percent_of_population_fully_vaccinated)
```

# Question 14: Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-11-15”?
```{r}
library(ggplot2)
```
```{r}
ggplot(tempSD, aes(percent_of_population_fully_vaccinated)) + geom_histogram()
```
## Focus on UCSD/La Jolla
```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```
# Question 15: Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area
```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x = "Date", y="Percent Vaccinated")
```
## Comparing to similar sized areas
```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-11-15")

#head(vax.36)
```
# Question 16: Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?
```{r}
mean(vax.36$percent_of_population_fully_vaccinated)
```
```{r}
#add geom_hline()
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  geom_hline(yintercept = 0.7172851) +
  ylim(c(0,1)) +
  labs(x = "Date", y="Percent Vaccinated")
```
# Question 17: What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”?
Min: 0.3785010
1st Qu.: 0.6396185
Median: 0.7155240
Mean: 0.7172851
3rd Qu.: 0.7879820
Max: 1.000000
```{r}
#min, 1st quarter, median, 3rd quarter, max
fivenum(vax.36$percent_of_population_fully_vaccinated)
#mean 
mean(vax.36$percent_of_population_fully_vaccinated)
```
# Question 18: Using ggplot generate a histogram of this data.
```{r}
ggplot(vax.36, aes(percent_of_population_fully_vaccinated)) + geom_histogram() +
  labs(x = "Percent Vaccinated", y = "Count")
```
# Question 19: Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?
Below, 92040 is 0.5467 and 92109 is 0.6933
```{r}
#92040
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
#92109
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```
# Question 20: Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.
```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color = "blue") +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination rate across California",
       subtitle="Only areas with a population above 36k are shown") +
  geom_hline(yintercept = 0.7172851, linetype = "dashed")
```
# Question 21: How do you feel about traveling for Thanksgiving Break and meeting for in-person class afterwards?
I am excited to travel for Thanksgiving Break and looking forward to meeting for in-person class afterwards.
