---
title: "Class 19: Pertussis Resurgence Mini Project"
author: "Garrett Cole"
format: gfm
---

## Is Pertussis on the ride?

The CDC track reported Pertussis cases in US and make their data available:
https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

```{r}
#Paste data using "Post as DF" from Addins
cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L),
  Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617)
)
```

# Question 1: With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.
```{r}
library(ggplot2)
```
```{r}
baseplot <- ggplot(cdc) + 
  aes(x = Year, y = Cases) +
  geom_point() +
  geom_line() +
  labs(title = "Number of reported Petussis Cases per Year (1922-2019)")
```
# Question 2: Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?
```{r}
baseplot + geom_vline(xintercept = 1946, linetype = "dashed", color = "blue") + geom_vline(xintercept = 1996, linetype = "dashed", color = "red")
```
# Question 3:  Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?
After the introduction of the aP vaccine, the number of cases started to slightly rise and had a minor jump but then are starting to stable out. One possible explanation is that the aP vaccine could cause the virus to mutate more rapidly causing new variants to come out.

## Getting Data from CMI-PB
The CMI-PB resource is studying and making available data on the immune response to Pertussis Vaccination

It mostly returns JSON format data that we need to process and convert into something usable in R.

```{r}
# Allows us to read, write and process JSON data
library(jsonlite)
```
```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
```
```{r}
head(subject, 3)
```
# Question 4: How may aP and wP infancy vaccinated subjects are in the dataset?
aP: 47, wP: 49
```{r}
table(subject$infancy_vac)
```
# Question 5: How many Male and Female subjects/patients are in the dataset?
66 Females and 30 Males
```{r}
table(subject$biological_sex)
```
# Question 6: What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?
```{r}
table(subject$race, subject$biological_sex)
```
## Working with dates

To help ease the pain working with dates we can use the library(lubridate)
```{r}
library(lubridate)
```
```{r}
today()
```
```{r}
# How many dates have passed since new year 2000
today() - ymd("2000-01-01")
```
```{r}
# Convert to years since new year 2000
time_length( today() - ymd("2000-01-01"),  "years")
```
# Question 7: Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?
```{r}
# Use todays date to calculate age in days
subject$age <- today() - ymd(subject$year_of_birth)
```
```{r}
library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )
```
```{r}
# wP
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```
# Question 8: Determine the age of all individuals at time of boost?
```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```
# Question 9: With the help of a faceted boxplot (see below), do you think these two groups are significantly different?
Yes, I think so since there is no overlap
```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) 
```
## Joining multiple tables
```{r}
# Complete the API URLs...
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/ab_titer", simplifyVector = TRUE) 
```

# Question 9:  Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:
```{r}
meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```
# Question 10: Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.
```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```
# Question 11: How many specimens (i.e. entries in abdata) do we have for each isotype?
```{r}
table(abdata$isotype)
```
# Question 12: What do you notice about the number of visit 8 specimens compared to other visits?
The number of visit 8 specimens is great amount less than the other visit specimens
```{r}
table(abdata$visit)
```

## Examine IgG1 Ab titer levels
```{r}
ig1 <- abdata %>% filter(isotype == "IgG1", visit!=8)
head(ig1)
```
# Question 13: Complete the following code to make a summary boxplot of Ab titer levels for all antigens:
```{r}
ggplot(ig1) +
  aes(MFI, antigen) +
  geom_boxplot() + 
  facet_wrap(vars(visit), nrow=2)
```
# Question 14: What antigens show differences in the level of IgG1 antibody titers recognizing them over time? Why these and not others?
TT, PTM, PT, PRN, FIM 2/3, and FHA show differences in the level of IgG1 antibody titers over time. I think maybe these and not the others is because they're the only ones with a MFI_normalised great than 1e+00
